---
title: "Analysis of Gun Violence in Schools"
subtitle: "Across the United States"
author: "Morgan Fisher"
format:
  revealjs:
    theme: solarized
    chalkboard: true
editor: visual
code-fold: true
cache: true
---


## Questions about the data

Throughout the project we will examine:

-   Has gun violence in schools increased over the past 30 years?
-   Are there states that experience more gun violence in schools?
-   What characteristics are common in schools that experience gun violence?


## Incidents Throughout Time

By graphing the number of incidents over time, we see large fluctuations from 1990 to 2020. However, ever since 2020, the number of incidents has largely increased with the most incidents ever recorded in 2023.
```{r}
library(tidyverse)
library(janitor)
library(dplyr)
library(ggplot2)


# read data
project_data <- readr::read_csv("pah_wikp_combo.csv", show_col_types = FALSE)

# clean up variable names
project_data <- janitor::clean_names(project_data)

# filter out duplicates
project_data <- project_data |>
  filter(is.na(dupe))

# remove unwanted variables & add year as variable
project_data <- project_data |>
  select(date, city, state, area_type, school, fatalities, wounded) |>
  mutate(year = as.numeric(substr(date, nchar(date) - 3, nchar(date))))

data_over_time <- project_data |>
  select(year) |>
  filter(year != 2024) |>
  group_by(year) |>
  summarise(count = n()) |>
  ungroup()

ggplot(data_over_time, aes(x = year, y = count)) +
  geom_line(color = "blue") +
  geom_point(color = "blue") +
  labs(x = "Year", y = "Number of Occurrences", title = "Occurrences Over Time (1990-2023)") +
  theme_minimal()

```

## Occurrences by State

By creating a map of occurrence rate for the US states, we see that occurrences are more likely in Southern states as well as New Mexico and Delaware

```{r}
library(maps)
library(sf)
library(ggplot2)
library(readxl)

# map of states
us_states <- st_as_sf(map("state", plot = FALSE, fill = TRUE))


# population data
population_data <- read_excel("apportionment-2020-table01.xlsx", skip = 3)

colnames(population_data) <- c("state", "population")

population_data <- population_data |>
  select(state, population)

occurrences_by_state <- project_data |>
   filter(!(state %in% c("D.C.", "District of Columbia", "District Of Columbia"))) |>
  group_by(state) |>
  summarise(total_occurrences = n()) |>
  ungroup()

occurrences_and_population <- left_join(occurrences_by_state, population_data, by = c("state" = "state"))

occurrences_and_population  <- occurrences_and_population |>
  mutate(occurrence_rate = total_occurrences / (population / 1000000)) |>
  mutate(state = tolower(state))

#fatalities_and_population

occurrences_us_map <- left_join(us_states, occurrences_and_population, by = c("ID" = "state"))

ggplot() +
  geom_sf(data = occurrences_us_map, aes(fill = occurrence_rate)) +
  labs(title = "Occurrence Rate by State (1990 to 2024)") +
  scale_fill_distiller(name = "Occurrence Rate", direction = 1) +
  theme_minimal()

```

## Incidents by School Type

High schools lead in incidents, followed by colleges, middle schools, and elementary schools. Fatalities are higher on average at elementary schools and colleges compared to middle schools and high schools.
```{r}
# clean up the data by removing the unknown values and renaming the other variables
school_type_data <- project_data |>
  filter(!is.na(school)) |>
  filter(!grepl("-", school, fixed = TRUE)) |>
  mutate(school = case_when(
    school == "C" ~ "College",
    school == "HS" ~ "High School",
    school == "MS" ~ "Middle School",
    school == "ES" ~ "Elementary",
    TRUE ~ as.character(school) ))

school_order <- c("Elementary", "Middle School", "High School", "College")

school_type_data <- school_type_data |>
  mutate(school = factor(school, levels = school_order))


school_type_table <- school_type_data |>
  select(school) |>
  group_by(school)|>
  summarise(count = n()) |>
  ungroup()

par(mfrow = c(1, 2), mar = c(1, 3, 3, 3))  # Set bottom margin to 2
#par(mar = c(1, 1, 1, 1))

percentages = round(school_type_table$count / sum(school_type_table$count) * 100)

#par(mfrow = c(1, 2), mar = c(1, 1, 1, 1))

piechart_1 <- pie(school_type_table$count,
    labels = paste0(school_type_table$school, "\n", paste0("(", percentages, "%)")),
    main = "Occurrences by School Type", cex = 0.8)

school_type_table2 <- school_type_data |>
  select(school, fatalities) |>
  group_by(school) |>
  summarise(fatalities_count = sum(fatalities)) |>
  ungroup()

#par(mar = c(1, 1, 1, 1))

percentages <- round(school_type_table2$fatalities_count / sum(school_type_table2$fatalities_count) * 100)


piechart_2 <- pie(school_type_table2$fatalities_count,
    labels = paste0(school_type_table2$school, "\n", paste0("(", percentages, "%)")),
    main = "Fatalities by School Type", cex = 0.8)


```

## Summary

From our data, we drew the following conclusions:

-   Large increase in school shootings from 2020 to 2023
-   Southern states as well as New Mexico and Delaware experience increased incidents compared to other states
-   The majority of incidents occur in high schools, but on average, incidents at colleges and elementary schools are more deadly

Looking at these conclusions, we can start to ask questions about why these patterns are occurring and how we can use these patterns to prevent further incidents in the future.




