---
title: "Graphing Relationships"
author: "Morgan Fisher"
format: html
editor: visual
code-fold: true
cache: true
warning: false
---

## Question 1

**Posted here is a data file containing presidential Democratic votes shares of the two party vote, by states (including DC), from 1980 to 2016, leipvote1980_2016wide.dta. Import that data (which is in Stata format) into R using haven and inspect it. The years run from 1980, 1984, …. , 2012, 2016.**

Importing the data:

```{r}
library(haven)

# Replace "your_file_path.dta" with the path to your .dta file
voter_data1980_2016 <- read_dta("leipvote1980_2016wide.dta")


# read data
voter_data2020 <- readr::read_csv("us_vote_2020.csv")

```

**1.  Left join the 2020 election data included (“us_vote_2020.csv”) to the 1980-2016 data frame. It is also posted on the assignment page. Create a variable that measures the percent of the two pary vote won by Biden in 2020 using the dem_votes and rep_votes variables and a mutate statement.**

```{r}
library(dplyr)

# joining data
voter_data1980_2020 <- left_join(voter_data1980_2016, voter_data2020, by = c("state" = "state"))

voter_data1980_2020 <- voter_data1980_2020 |>
  #creating a new variable
  mutate(pctdem2020 = (dem_percent / (dem_percent + rep_percent))) |>
  #removing unwanted variables
  select(-called, -final, -dem_votes, -rep_votes, -other_votes, -dem_percent, -rep_percent, -other_percent, -dem_this_margin, -margin_shift, -vote_change, -stateid)

  
head(voter_data1980_2020, 5)
```

**2.  Create a table of summary statistics (means, standard deviations, and number of states) for the 2016 and 2020 Democratic vote shares. Describe the overall pattern, in one sentence.**

```{r}
library(tidyr)
library(dplyr)
library(gt)


voter_data_tbl <- voter_data1980_2020|>
  #selecting specific columns
  select(state, pctdem2016, pctdem2020)|>
  #summarizing by mean, sd, & count
  summarise(mean_2016 = mean(pctdem2016), 
            sd_2016 = sd(pctdem2016),
            count_2016 = n(),
            mean_2020 = mean(pctdem2020), 
            sd_2020 = sd(pctdem2020),
            count_2020 = n()
            ) |>
  #pivting longer
  pivot_longer(cols = everything(), names_to = "Statistic", values_to = "Value") |>
  mutate(Year = ifelse(grepl("_2016", Statistic), "2016", "2020")) |>
  separate(Statistic, into = c("Statistic", "Year"), sep = "_") |>
  pivot_wider(names_from = Statistic, values_from = Value)

gt_tbl <- gt(voter_data_tbl) |>
  #creating title
  tab_header(
    title = "2016 and 2020 Democratic Vote Share Statistics",
    subtitle = "Mean, Standard Deviation, and Count of States") |>
  #editing labels
  cols_label(
    Year = "Year",
    mean = "Mean",
    sd = "Standard Dev.",
    count = "Count")

gt_tbl
```

Compared to the 2020 election, the 2016 election had a lower percentage of Demoratic votes but higher variance between states.


## Question 2 

**Using the data from 1, generate a vote swing scatter plot with the 2020 Democratic percentage of the vote on the Y axis and the 2016 Democratic percentage of the vote on the X axis, labeling the points with the state names, and adding a 45 degree line to the plot using geom_abline().**

```{r}
library(ggplot2)
library(ggrepel)

ggplot(data = voter_data1980_2020,
            mapping = aes(x = pctdem2016, y = pctdem2020, label = state)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "blue") +
  geom_text_repel(size = 2, nudge_x = 0, nudge_y = 0.1) +
  labs(x = "Democratic Percentage in 2016", y = "Democratic Percentage in 2020", title = "Vote Swing Scatter Plot") +   theme_minimal()
```

**1.  Describe your graph. What is is overall pattern in the data?**

There is a strong correlation between Democratic Percentage in 2016 & 2020. Also, most states have a higher democratic percentage in 2020 than they did in 2016.

**2.  What challenges are created by labeling the states? Try labeling only some of the states, justifying your choice of states.**

Because there are so many states and the data points are so close together, it makes it difficult to clearly label each state without overlap.

**3.  Are there unusual states in the plot? If so, which ones?**

Yes, District of Columbia has a far higher Democratic percentage in both 2016 & 2020 compared to all other states.

**4.  Make a graph that omits Washington D.C. from the data. What effect does that have on the structure of the graph?**

```{r}
library(ggplot2)
library(ggrepel)

data_no_dc <- voter_data1980_2020 |>
  filter(state != "District of Columbia")

ggplot(data = data_no_dc,
            mapping = aes(x = pctdem2016, y = pctdem2020, label = state)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "blue") +
  geom_text_repel(size = 2, nudge_x = 0, nudge_y = 0.1) +
  labs(x = "Democratic Percentage in 2016", y = "Democratic Percentage in 2020", title = "Vote Swing Scatter Plot w/o DC") +   theme_minimal()
```

Ommiting Washington DC allows the graph to have a narrower window, and we can now see the difference of each state more prominently.





## Question 3 

**Again using the data from 1, filter out Washington D.C., reshape the data from wide form to long form using pivot_longer() and generate a line plot for the 50 states from 1980 to 2020, graphing percent Democratic on the y-axis and year on the x-axis. Choose a set of states that you find interesting, and emphasize those states using colors. \[Warning: the initial plots will be a bit messy because you are plotting 50 states.\]**

```{r}
data_no_dc_long <- data_no_dc |> pivot_longer(cols = starts_with("pctdem"),
               names_to = "year",
               values_to = "percentage_democratic",
               names_pattern = "pctdem(.*)")


state_colors <- c("Wyoming" = "purple", "Vermont" = "red", "Georgia" = "magenta", "Utah" = "turquoise")

highlight_states <- c("Wyoming", "California", "Vermont", "West Virginia")
data_highlighted <- data_no_dc_long %>% filter(state %in% highlight_states)


ggplot(data = data_no_dc_long,
            mapping = aes(x = year, y = percentage_democratic, group = state, color = state)) +
  geom_line(size = 0.5) +
  scale_color_manual(values = state_colors) +
  labs(x = "Year", y = "Democratic Percentage", title = "Democratic Percentage from 1980 to 2020") +  
  theme_minimal() +
  geom_line(data = data_highlighted, aes(group = state, color = state), size = 0.5)
  
```


**1.  Describe the overall patterns in the data over time. What central features are particularly noteworthy, and why?**

States appear to have more variation from one another in their democratic percentages as time goes on.

**2.  What story are you telling with your use of color to highlight a set of states?**

I decided to highlight the states with the highest & lowest democratic percentage in both 1980 & 2020. This shows how a state demographics can change over time. We can see that Vermont and Utah favored Democrats more from 1980 to 2020, while Georgia and Wyoming favored Democrats less.

**3.  Remember that in section 7.1, Healy faceted his election graph by the four census regions. Generate a second graph structured as before, this time faceting by census region.**

```{r}
library(dplyr)
library(ggplot2)

midwest_states <- c("Illinois", "Indiana", "Iowa", "Kansas", "Michigan", "Minnesota", "Missouri", "Nebraska", "North Dakota", "Ohio", "South Dakota", "Wisconsin")

northeast_states <- c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont","New Jersey", "New York", "Pennsylvania")

south_states <- c("Alabama", "Arkansas", "Delaware", "Florida", "Georgia", "Kentucky", "Louisiana", "Maryland", "Mississippi", "North Carolina", "Oklahoma", "South Carolina", "Tennessee", "Texas", "Virginia", "West Virginia")

west_states <- c("Alaska", "Arizona", "California", "Colorado", "Hawaii", "Idaho", "Montana", "Nevada", "New Mexico", "Oregon", "Utah", "Washington", "Wyoming")

data_midwest <- data_no_dc_long %>%
  filter(state %in% midwest_states)

data_south <- data_no_dc_long %>%
  filter(state %in% south_states)

data_west <- data_no_dc_long %>%
  filter(state %in% west_states)

data_northeast <- data_no_dc_long %>%
  filter(state %in% northeast_states)

combined_data <- bind_rows(
  data_midwest %>% mutate(region = "Midwest"),
  data_northeast %>% mutate(region = "Northeast"),
  data_south %>% mutate(region = "South"),
  data_west %>% mutate(region = "West")
)

# Plotting all regions faceted
ggplot(data = combined_data,
       aes(x = year, y = percentage_democratic, group = state)) +
  geom_line(size = 0.5) +
  labs(x = "Year", y = "Democratic Percentage", title = "Democratic Percentage from 1980 to 2020") +
  facet_wrap(~ region, nrow = 2) +
  theme_minimal()  +
  theme(axis.text.x = element_text(size = 6))
```
